Towards a reproducible framework for undergraduate data science research

### 2019-11-30 Rmd/introduction.Rmd
fixed headers
updated methods

## Introduction

@@

@nolan2010computing argue for students to work with real data. 
Working with real data allows them to develop skill not only in statistical analysis,
but also in data transfer from online sources, in data storage, and in using data 
from multiple file formats. In the case of Twitter data, tweets are stored in
Javascript Object Notation (JSON) [@tweet_json, @json].

Mentoring in the work place and in higher education can have many benefits, including [@higgins2001reconceptualizing].
A key component of our mentoring framework is the emphasis on using real data to answer real 
scientific questions. We believe that this process develops the problem-solving skills that 
students will need in their future careers in data science. 
We encouraged the student to articulate a scientific research question, translate that question 
into quantitative and statistical terms, determine which data could be used to address the 
question, acquire the data, analyze data, visualize results, and communicate what they learned.

@@

During the time when we first implemented our framework, we served as early-career instructors in 
the statistics department at the University of Wisconsin-Madison. One of us (Hanlon) had prior 
experience in mentoring undergraduates, while the other (Boehm) had none. Our initial 
conceptualization of mentoring drew heavily on ideas we first encountered in professional development 
courses, including the Delta Program's Entering Mentoring class (https://delta.wisc.edu). 
Professor Erik Nordheim heavily 
influenced our approach to and philosophy of teaching statistics. 
We studied with Professor Nordheim early in our teaching 
careers, and his guidance continues to influence our teaching practices.

We both have experience in teaching undergraduate introductory statistics courses with enrollments over 
100 students. Through our interactions with students in these classes, we've come to value not only 
the ideas in a traditional introductory course, but also the need to prepare students with the essential skills needed for success in data science. @nolan2010computing summarizes these skill sets in the following three ideas:

@@

4. develop skills in reproducible research to promote the practice of open science

We see the fourth goal as an equal with the first three from @nolan2010computing. 
Data scientists are uniquely positioned to promote open science practices, including the free sharing of 
data, code, and instructions for their use. The recently recognized need for science to be more transparent and more reproducible elevate this goal to the level of the first three.

>>>>

# Introduction

@@

@nolan2010computing argue for students to work with real data. 
Working with real data allows them to develop skill not only in statistical analysis,
but also in data transfer from online sources, in data storage, and in using data 
from multiple file formats. In the case of Twitter data, tweets are stored in
Javascript Object Notation (JSON) [@tweet_json,@json].

Mentoring in the work place and in higher education can have many 
benefits, including improving students' development as thinkers and scholars, confidence in their 
own abilities, integration into the campus community, and interest in graduate training 
[@baker2010beyond,@higgins2001reconceptualizing].
A key component of our data science mentoring framework is the emphasis on using real data to answer real 
scientific questions. We believe that this process develops problem-solving skills that 
students will need in their future careers in data science. 
We encouraged the student to articulate a scientific research question, translate that question 
into quantitative and statistical terms, determine which data could be used to address the 
question, acquire the data, analyze data, visualize results, and communicate what they learned.

@@

During the time when we first implemented our framework, we served as early-career instructors in 
the statistics department at the University of Wisconsin-Madison. One of us (Hanlon) had prior 
experience in mentoring students, while the other (Boehm) had none. Our initial 
conceptualization of mentoring drew heavily on ideas we first encountered in professional 
development 
activities, including the Delta Program's mentoring class (https://delta.wisc.edu) and @handelsman2005entering. 
Professor Erik Nordheim heavily 
influenced our approach to and philosophy of teaching statistics. 
We studied with Professor Nordheim early in our teaching 
careers, and his emphasis on active learning continues to influence our teaching practices.

We both have experience in teaching undergraduate introductory 
statistics courses with enrollments over 
100 students. Through our interactions with students in these 
classes, we've grown to value not only 
the ideas in a traditional introductory course, but also the need 
to prepare students with the essential skills needed for success 
in data science. @nolan2010computing summarizes these skill sets 
in the following three goals:

@@

4. develop skills in reproducible research to promote open science practices

We see the fourth goal as an equal with the first three from @nolan2010computing. 
Data scientists are uniquely positioned to promote open science 
practices, including the free sharing of 
data, code, and instructions for their use. 
The need for science to be more transparent and more reproducible
elevate this goal to the level of the first three.

### 2019-11-30 Rmd/methods.Rmd
updated methods

We designed and used a framework for mentored undergraduate data science research 
projects with big data. Our framework draws heavily on ideas in @nolan2010computing. 
Specifically, we incorporate the three key aspects that @nolan2010computing identified:

@@

Second, the students worked with a variety of data structures. 
The Twitter streaming 
API returns tweets as JSON (Javascript Object Notation, http://json.org). Because 
distinct Twitter users may provide 
different pieces of profile information, there is variability in the structure of 
each tweet's JSON. 
Additionally, tweet metadata fields may appear in any order (https://developer.twitter.com/en/docs/tutorials/consuming-streaming-data). 
Students needed to recognize this and to write code that 
accommodated these variations in tweet data structure.

Students wrote code to parse and organize tweet JSONs. They organized their R code into a package, https://github.com/rturn/parseTweetFiles. 
Each tweet's JSON included required fields, and, possibly, some optional fields. 
The required fields are:

**see https://developer.twitter.com/en/docs/tweets/data-dictionary/overview/tweet-object**
**insert a 2-column table of fields here**

The evolving nature of JSON tweet structure 
(https://developer.twitter.com/en/docs/tweets/data-dictionary/guides/tweet-timeline) 
required students to write flexible code that could incorporate newly introduced 
or deprecated metadata.

@@

Both drew heavily on our collection of tweets. One student examined Standard and Poor's 500 index daily 
closing prices over time. She also analyzed sentiments from each day's stock market-related tweets 
to look for relationships between tweet sentiment and stock market prices.

@@

## Implementation of the framework

>>>>

We designed and used a framework for mentored undergraduate data science research 
projects with big data. Below, we describe our initial implementation of the framework before relating 
it to three major ideas from @nolan2010computing.

## Framework implementation

## Relating to three ideas from @nolan2010computing

We incorporate three key aspects that @nolan2010computing identified:

@@

Second, the students worked with a variety of data structures. 
The Twitter streaming 
API returns tweets as JSON (Javascript Object Notation). Because 
distinct Twitter users may provide 
different pieces of profile information, there is variability in the structure of 
each tweet's JSON. 
Additionally, tweet metadata fields may appear in any order (https://developer.twitter.com/en/docs/tutorials/consuming-streaming-data). 
Students needed to recognize this and to write code that 
accommodated these variations in tweet data structure.
Additional variability in tweet structure arose due to changes in the API. 
The evolving nature of JSON tweet structure 
(https://developer.twitter.com/en/docs/tweets/data-dictionary/guides/tweet-timeline) 
required students to write flexible code that could incorporate newly introduced 
or deprecated metadata.

Students wrote R code to parse and organize tweet JSON []. They organized their R code into a package, and shared it on Github (https://github.com/rturn/parseTweetFiles). 
Each tweet's JSON included required fields, and, possibly, some optional fields.

@@

Both drew heavily on the collection of tweets. One student examined 
Standard and Poor's 500 index daily 
closing prices over time. She also analyzed sentiments 
from each day's stock market-related tweets 
to look for relationships between tweet sentiment and stock market prices.

@@

Both students analyzed tweets as texts. This first required them to write code to parse the JSON
that the API returns. Once they had isolated the tweet text from its metadata, they parsed the 
tweet text into words for use in sentiment analysis and topic modeling. 
For the stock market project, they analyzed only those tweets that contained finance-related 
keywords. Sentiment analysis involved comparisons of tweet words to a dictionary that mapped words to 
sentiments. This yielded a net sentiment score for each tweet. 
They then treated tweet sentiment scores as a time series and compared them with 
daily stock market index closing prices.

The second student project involved latent Dirichlet allocation modeling of
tweet words at distinct time points to detect social media events [@blei2003latent]. 
Latent Dirichlet allocation is a bayesian nonparametric method for modeling text corpora as the result 
of words chosen from topics.
